Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-01 frontiers/frontiers.Rnw
Added detail to background of frontiers paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@228 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function
(though extension to other discrepancy functions is straightforward).
The tests are carried out after model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
``posthoc'' test to examine whether certain parameters instead vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, these statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $l(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  In words, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  Scores close to zero
indicate a ``good'' description, and scores far from zero indicate a
``bad'' description.  Each of an individual's $k$ scores represent one
model parameter, so that the scores provide specific information about
each parameter's fit to each individual.

%% Next: How scores can be used for testing

>>>>

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being straightforward.
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $l(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
of these scores forms the ``meat'' for the calculation of robust
(Huber-White) standard errors. %TODO cite

To expand on the above equation, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  Scores close to zero
indicate a ``good'' description, and scores far from zero indicate a
``bad'' description.  Each of an individual's $k$ scores represent one
model parameter, so that the scores provide specific information about
each parameter's fit to each individual.

To use these scores for testing, we order individuals according to
an auxiliary variable (the variable against which we are testing 
measurement invariance) and look for trends in the
scores (roughly speaking).  For example, imagine that we are testing
for measurement 
invariance w.r.t.\ age.  If the manifest variables violate measurement
invariance here, then some parameter estimates may be too large for
young individuals and too small for old individuals (say).  This
result would be reflected in the scores, where young individuals'
scores would be greater than zero and old individuals' scores would be
less than zero.  Conversely, if measurement invariance holds, then all
individuals' scores will fluctuate around zero.

To formalize the ideas in the previous paragraph, we focus on a cumulative
sum of the ordered scores.  Specifically, we focus on how the
cumulative sum fluctuates as more individuals' scores are added to it,
starting with the youngest and ending with the oldest individual
(assuming that age is the auxiliary variable of interest).  Under the
hypothesis of measurement invariance, a central limit theorem can be
used to show that the fluctuation of the cumulative sum follows a
Brownian bridge.  This result allows us to calculate $p$-values and
critical values for test statistics under the hypothesis of
measurement invariance.  This includes test statistics associated with
all model parameters and subsets of model parameters.

%% TODO focus on ordinal
Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores.  One could
take the absolute maximum that the cumulative sum attains for any
parameter of interest, resulting in a {\em double max} statistic.
Alternatively, one could sum the (squared) cumulative sum across parameters
of interest and take the maximum across individuals, resulting in a
{\em maximum Lagrange multiplier} statistic 
\cite<see>[for further discussion of this idea]{MerZei13}.
